A Type System for String Sanitation\\Implemented Inside a Python

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Fragment -> full system, but the rules don't fit on a single page so we might have to revert (or else get fancy with space)
Fixed typos due to out-of-date \dosubst macro
Factored s \in \lang{r} out of the theorem previously numbered as 10.
This property is now a property of \lambdas (thm 9) which is preserved under
translation (them 13). In order to prove them 13, we need thm 12 which is
basically CF + thm 11

@@

% Meta-theoretic functions
\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}

@@

<$\psi$> ::=	...			\hfill	source types					\alt
$\stringin{r}$

<S> ::= ... \hfill source terms \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{S}{S}$ \alt
      $\sisubst{r}{S}{S}$ \alt
      $\coerce{r}{S}$
\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
\end{grammar}
\end{figure}

@@

<$\theta$> ::= ... \hfill target types \alt
$\str$ \alt $\regex$

<P> ::= ... \hfill target terms \alt
  $\tstr{s}$ \alt
  $\rx{r}$ \alt
  $\tconcat{P}{P}$ \alt
  $\preplace{P}{P}{P}$ \alt
  $\tcheck{P}{P}$ 
  \end{grammar}
\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

@@

% The translations rules *just won't* go to the bottom of the page, so 
% instead I am placing them at the top. Since the figures are out of order,
% it's even more important that we get the number gith. therefore, the counter
% is modified before and after this figure.
\addtocounter{figure}{4}
\begin{figure*}\label{fig:trans}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
\begin{mathpar}
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-replace]
{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}

@@

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-Replace]
{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}

@@

\begin{mathpar}
\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\begin{mathpar}
\inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Replace]
{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}

@@

\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
The following are well-known properties of regular expressions which are necessary for our proofs:
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
Regular languages are closed under complements and concatenation.
The regular expressions correspond bijectively to the regular languages.
\end{lem}

@@

\begin{lem}
  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression. 
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
\end{proof}

\begin{lem}
  If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\end{lem}

@@

@@

We now present the main correctness result.

\begin{thm}
If $S : \strin{r}$ then there exists a P such that $\trden{s} = P$ and either: 
(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$; or
(b) $P \err$ and $S \err$.
\end{thm}

\begin{proof}
The proof proceeds by induction on the typing relation for $S$ and an appropriate
choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
by the standard error propagation rules omitted for space.
Throughout the proof, properties from the closure lemma for regular languages
are necessary.

%\begin{description}
%\item[S-T-String-I:] 
%Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
%Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
%By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
%Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
%
%\item[S-T-Concat:]
%Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
%It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
%By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
%Choose $P = \concat{P_1}{P_2}$.
%
%We first prove property (a).
%Note that $S_1$ and $P_1$ do not result in errors and that $S_1$ is well typed.
%Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
%Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
%Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
%Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
%
%Consider property (b).
%If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
%Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
%
%\item[S-T-Replace:]
%Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
%By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
%such that $\lsubst{r}{r_1}{r_2} = r'$.
%By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
%and either (a) or (b) holds.
%
%If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
%and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
%Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
%Choose $P = \preplace{r}{s_1}{s_2}$.
%By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
%What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
%which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
%
%If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
%Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
%
%\item[S-T-Coerce:]
%Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
%By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
%By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
%holds for $S'$ and $P'$.
%
%If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
%Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
%Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
%Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
%Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
%and S-E-Coerce-Err.
%
%Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
%
%\end{description}
\end{proof}

>>>>

\newcommand{\val}{{\sf val}}

@@

% Meta-theoretic functions
\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}

@@

<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
$\stringin{r}$

<S> ::= 
      $\lambda x . e$ \hfill source terms \alt
      $e e$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{S}{S}$ \alt
      $\sisubst{r}{S}{S}$ \alt
      $\coerce{r}{S}$
\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
\end{grammar}
\end{figure}

@@

<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
$\str$ \alt $\regex$

<P> ::= $\lambda x . e$ \hfill target terms \alt
  $e e$ \alt
  $\tstr{s}$ \alt
  $\rx{r}$ \alt
  $\tconcat{P}{P}$ \alt
  $\preplace{P}{P}{P}$ \alt
  $\tcheck{P}{P}$ 
  \end{grammar}
\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

@@

% The translations rules *just won't* go to the bottom of the page, so 
% instead I am placing them at the top. Since the figures are out of order,
% it's even more important that we get the number gith. therefore, the counter
% is modified before and after this figure.

\addtocounter{figure}{4}
\begin{figure*}\label{fig:trans}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
\begin{mathpar}
%\inferrule[S-T-Abs]
%{\sctx, x : T_1 \vdash e : T_2}
%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[S-T-App]
%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
%{\sctx \vdash e_1 e_2 : T}
%  
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-Replace]
{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}

@@

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
%
%\inferrule[S-E-Abs]                                                             
%{ \ }                                                                           
%{\lambda x.e \sreduces \lambda x.e}                                             
%
%\inferrule[S-E-App]
%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-Replace]
{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}

@@

\begin{mathpar}
%\inferrule[P-T-Abs]
%{\tctx, x : T_1 \vdash e : T_2}
%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[P-T-App]
%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
%{\tctx \vdash e_1 e_2 : T}
%
\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\begin{mathpar}

%\inferrule[P-E-Abs]
%{ \ }
%{\lambda x.e \sreduces \lambda x.e}
%
%\inferrule[P-E-App]
%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
  \inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Replace]
{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}

@@

\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
The following are properties of regular expressions which are necessary for our proofs:
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
\end{lem}

@@

\begin{lem}
  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
\end{proof}
\begin{lem}
If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\end{lem}

@@

In addition to safety, we proof a correctness result for $\lambdas$ which will
be preserved under translation.

\begin{thm}{Input Sanitation Correctness of $\lambdas$}\label{thm:scorrect}
  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
  or (b) $S \err$.
\end{thm}
\begin{proof}
  We may actually want a proof here.
\end{proof}

Finally, $\lambdap$ is also type-safe. We omit the proof due to space constraints,
and because the proof is trivial.

@@

\begin{thm}{Translation Correctness} \label{thm:tcorrect}
If $\sctx \vdash S : \stringin{r}$ then there exists a $P$ such that $\trden{S} = P$ and either: 
(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or
(b) $P \err$ and $S \err$.
\end{thm}

\begin{proof}
The proof proceeds by induction on the typing relation for $S$ and an appropriate
choice of $P$; in each case, the choice is obvious. The subcases (a) proceed
by inversion and appeals to our type safety theorems as well as the induction hypothesis. The subcases (b) proceed
by the standard error propagation rules omitted for space.
Throughout the proof, properties from the closure lemma for regular languages
are necessary.
\end{proof}

\begin{thm}
If If $\sctx \vdash S : \stringin{r}$ and $\trden{S} = P$ then either:
(a) $P \treduces \tstr{s}$ and $S \sreduces \strin{s}$, or                      
(b) $P \err$ and $S \err$. 
\end{thm}
\begin{proof}
  The theorem  follows directly from canonical forms for translations and theorem \ref{thm:tcorrect}.
\end{proof}

Finally, our main theorem establishes that input sanitation correctness of $\lambdas$ is preserved under the translation into $\lambdap$.

\begin{thm}{Input Sanitation Correctness for Translated Terms}
  If $\trden{S} = P$ and $\sctx \vdash S : \stringin{r}$ then either $P \err$ or $P \sreduces \tstr{s}$
  for $s \in \lang{r}$.
\end{thm}
\begin{proof}
  By theorem \ref{tcorrect}, $P \sreduces \tstr{s}$ implies that $S \sreduces \strin{s}$.
  By theorem \ref{scorrect}, this together with the assumption that $S$ is well-typed implies that $s \in \lang{r}$.
\end{proof}
% PROOF OF TRANSLATION CORRECTNESS.
% This proof is slightly out of say. It's SUFFICIENT, but is actually more
% than sufficient because we prove \ref{scorrect} along with translation
% correctness....
%\begin{description}
%\item[S-T-String-I:] 
%Let $S = \rstr{s}$ and suppose $\sctx \vdash \rstr{s} : \stringin{r}$.
%Choose $T = \str{s}$ and note that $\trden{S} = P$ by Tr-String.
%By P-E-String, $P \treduces \str{s}$ and by S-E-String $S \treduces \rstr{s}$.
%Finally, by inversion of S-T-Stringin-I, $s \in \lang{r}$.
%\item[S-T-Concat:]
%Let $S = \rconcat{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r_1 r_2}$.
%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$.
%It follows by induction that there exists a $P_1$ such that $\trden{S_1} = P_1$.
%By a similar argument for $S_2$ and $r_2$, there exists a $P_2$ such that $\trden{S_2} = P_2$.
%Choose $P = \concat{P_1}{P_2}$.
%We first prove property (a).
%Note that $S_1$ and $P_1$ do not result in errors and that $S_1$ is well typed.
%Therefore, $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$ by theorems 8 and 9 respectively. % TODO-nrf make sure this numbering is correct
%Similarly, $S_2 \sreduces \rstr{s_2}$ and $P_2 \treduces \str{s_2}$ for some $s_2 \in \lang{r_2}$.
%Therefore, $S \sreduces \rstr{s_1 s_2}$ by S-E-Concat and $\concat{P_1}{P_2} \treduces \str{s_1 s_2}$ by P-E-Concat.
%Finally, $s_1s_2 \in \lang{r_1}{r_2}$ by \ref{thm:regexprops}.
%Consider property (b).
%If $S_1 \err$ then $P_1 \err$ by induction, and it follows that $S \err$ and $P \err$ by respective error propagation rules.
%Similarly, if $S_2 \err$ then $P_2 \err$ and it follow that $S \err$ and $P \err$ by induction and propagation.
%\item[S-T-Replace:]
%Let $S = \rreplace{r}{S_1}{S_2}$ and suppose $\sctx \vdash S : \stringin{r'}$ for some $s$.
%By inversion of S-T-Replace, , $\sctx S_1 : \stringin{r_1}$ and $\sctx : \stringin{r_2}$ 
%such that $\lsubst{r}{r_1}{r_2} = r'$.
%By induction, there exists some $P_1, P_2$ such that $\trden{S_1} = P_1$, $\trden{S_2} = P_2$
%and either (a) or (b) holds.
%If (a) holds then $S_1 \sreduces \rstr{s_1}$ and $P_1 \treduces \str{s_1}$ for some $s_1 \in \lang{r_1}$,
%and similarly for $S_2, P_2$ and some $s_2 \in \lang{r_2}$.
%Therefore, by S-E-Replace, $S \sreduces \rstr{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
%Choose $P = \preplace{r}{s_1}{s_2}$.
%By a similar argument and P-E-Replace, $P \treduces \str{s}$ for some $s = \lreplace{r}{s_1}{s_2}$.
%What remains to be shown is $\lreplace{r}{s_1}{s_2} \in \lang{\lsubst{r}{r_1}{r_2}}$,
%which follows from Leamm D since $s_1 \in r_1$ and $s_2 \in r_2$.
%If (b) holds for $S_1$ and $P_1$, then $S \err$ and $P \err$ by propagation rules.
%Similarly, if (b) hols for $S_2$ and $P_2$ then $S \err$ and $P \err$ by propagation rules.
%\item[S-T-Coerce:]
%Let $S = \rcoerce{r}{S'}$ and suppose $\sctx \vdash \rcoerce{r}{S} : \stringin{r}$.
%By inverstion $\Psi \vdash S' : \stringin{r'}$ for an arbitrary $r'$.
%By induction there exists a $P'$ such that $\trden{S'} = P'$ and either (a) or (b)
%holds for $S'$ and $P'$.
%If (a) holds then $S' \sreduces \rstr{s'}$ and $P' \treduces \str{s'}$ for some $s' \in \lang{r'}$.
%Note that either $s' \in \lang{r}$ or $s' \not \in \lang{r}$ by property 2 of \ref{thm:regexprops}.
%Suppose $s' \in \lang{r}$. Then $\rcoerce{r}{S} \sreduces \rstr{S'}$ by S-E-Coerce.
%Choose $P = \check{\rx{r}}{P'}$ and note that $P \treduces \str{s'}$ by P-E-Coerce.
%Now suppose $s' \not \in \lang{r}$. Then $S \err$ and $P \err$ by P-E-Check-Err
%and S-E-Coerce-Err.
%Finally, if (b) holds then $S \err$ and $P \err$ by propagation.
%\end{description}
